---
title: "EDA Sandbox CLK"
author: "Christine Lucille Kuryla"
date: "2024-11-25"
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

This document is where Christine will play around with random stuff for funsies. 

# Article and Data

# Article

Article:
Hou, N., Li, M., He, L. et al. Predicting 30-days mortality for MIMIC-III patients with sepsis-3: a machine learning approach using XGboost. J Transl Med 18, 462 (2020). https://doi.org/10.1186/s12967-020-02620-5

Article's data subset of MIMIC-III:
https://translational-medicine.biomedcentral.com/articles/10.1186/s12967-020-02620-5 

## Load data

```{r}
# Downloaded from: https://translational-medicine.biomedcentral.com/articles/10.1186/s12967-020-02620-5#Sec14

# This data has all of the patients included in the study, after they filtered out certain patients due to the exclusion criteria in the paper
# n = 4559

data <- read_csv("data/12967_2020_2620_MOESM1_ESM.csv")

```

# Corr matrix heatmap

# PCA

## Numerica data only 

```{r}
data_numeric <-  data %>%
   # select(-c(icustay_id...1, hadm_id...2, icustay_id...103, hadm_id...102,intime, outtime, dbsource, blood_culture_time, antibiotic_time_poe, suspected_infection_time_poe)) %>% 
 #   select(where(is.numeric))  %>% 
   select(suspected_infection_time_poe_days,
          age, 
          icu_los,
          hosp_los,
          sofa,
          lods,
          sirs,
          qsofa,
          aniongap_min:glucose_mean,
          urineoutput,
          icustay_id...1,
          thirtyday_expire_flag) %>% 
  na.omit()
```

## PCA with prcomp

```{r}
pca_result <- prcomp(data_numeric %>% select(-c(icustay_id...1),
                                             thirtyday_expire_flag), scale = TRUE)

loadings <- pca_result$rotation

loadings <- loadings[,1:10]
```

### Loadings heatmap

```{r}
# Convert the loadings matrix into a long format for ggplot
library(reshape2)
loadings_melted <- melt(loadings)

# Plot the heatmap using ggplot
ggplot(loadings_melted, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "lightyellow", high = "red", midpoint = 0) +
  labs(title = "Heatmap of PCA Loadings", x = "Principal Components", y = "Variables") +
  theme_minimal()
```

### Biplot

```{r}
biplot(pca_result)

# library("FactoMineR")
# 
# fviz_pca_var(pca_result, col.var = "cos2",
#             gradient.cols = c("black", "orange", "green"),
#             repel = TRUE)
```

### PCA Scatterplot

```{r}
# Prepare data for plotting
scores <- as.data.frame(pca_result$x[, 1:6])  # Take PC1 and PC2
scores$thirty_day <- data_numeric %>% mutate(thirtyday_expire_flag = as.factor(thirtyday_expire_flag)) %>% pull(thirtyday_expire_flag) 
#colnames(scores) <- c("PC1", "PC2", "PC3", "PC4", "PC5", "thirty_day")

# Plot
ggplot(scores, aes(x = PC1, y = PC2, color = thirty_day)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(title = "PC1 vs PC2", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() 

as.data.frame(loadings) %>%
  select(PC1) %>% 
  arrange(PC1)


# Plot
ggplot(scores, aes(x = PC3, y = PC4, color = thirty_day)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(title = "PC1 vs PC2", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() 

# Plot
ggplot(scores, aes(x = PC2, y = PC5, color = thirty_day)) +
  geom_point(size = 2, alpha = 0.6) +
  labs(title = "PC1 vs PC2", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal() 


```

# Correlations

## Correlations of numeric variables

```{r}

correlation_matrix <- cor(data_numeric)
knitr::kable(correlation_matrix)

# Create a heatmap

#upper_tri <- matrixcalc::upper.triangle(correlation_auto)
#melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)

melted_cormat <- reshape2::melt(correlation_matrix, na.rm = TRUE)

ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
     hjust = 1))+
 coord_fixed() #+
 # geom_text(aes(Var2, Var1, label = if_else(value != 0, as.character(round(value, digits = 2)), " ")))


```

### Interactive 

```{r}

# Load necessary packages
library(ggplot2)
library(reshape2)
library(plotly)
library(Hmisc)


# Compute the correlation matrix and p-values
cor_res <- rcorr(as.matrix(data_numeric))

cor_mat <- cor_res$r        # Correlation coefficients
p_mat <- cor_res$P          # P-values

# Melt the correlation matrix and p-value matrix into long format
cor_mat_melt <- melt(cor_mat)
p_mat_melt <- melt(p_mat)

# Rename columns for clarity
colnames(cor_mat_melt) <- c("Var1", "Var2", "Correlation")
colnames(p_mat_melt) <- c("Var1", "Var2", "p_value")

# Combine the correlation and p-value data
cor_mat_melt$p_value <- p_mat_melt$p_value

# Create a column for significance asterisks
cor_mat_melt$signif <- ifelse(cor_mat_melt$p_value < 0.05, "*", "")

# Create a ggplot heatmap with custom text for tooltips
p <- ggplot(cor_mat_melt, aes(
    x = Var1,
    y = Var2,
    fill = Correlation,
    text = paste0(
        "Variables: ", Var1, " & ", Var2, "<br>",
        "Correlation: ", round(Correlation, 2), "<br>",
        "p-value: ", signif(p_value, 2)
    )
)) +
    geom_tile() +
    geom_text(aes(label = signif), color = "black", size = 1) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    )

# Convert ggplot object to an interactive Plotly object
p_interactive <- ggplotly(p, tooltip = "text")

# Display the interactive heatmap
p_interactive


```


